~\Large Increased sexual dimorphism evolves in a fossil stickleback following ecological release from fish piscivores

### 2024-07-25 manuscript/paper.tex
manuscript additions

Ecological release theory suggests that a prey population's niche should
expand when predation is relaxed or removed, as prey take advantage of
new enemy free space (reviewed in Herrmann, Stroud, and Losos
(\protect\hyperlink{ref-Herrmann2021}{2021})). Niche expansion could
manifest through a combination of increasing within- and
between-individual niche widths (Bolnick et al.
(\protect\hyperlink{ref-Bolnick2010}{2010}); Herrmann, Stroud, and Losos
(\protect\hyperlink{ref-Herrmann2021}{2021})), including divergence
between the sexes (Bolnick and Doebeli
(\protect\hyperlink{ref-Bolnick2003}{2003}); Cooper, Gilman, and
Boughman (\protect\hyperlink{ref-Cooper2011}{2011})). Theory and
empirical data suggest that a population presented with ecological
opportunity following ecological release might experience disruptive
selection on males and females stemming from intraspecific competition
over newly accessible resources, resulting in intersexual divergence in
habitat use (W. (\protect\hyperlink{ref-Schoener1968}{1968}); Shine
(\protect\hyperlink{ref-Shine1989}{1989}); Bolnick and Doebeli
(\protect\hyperlink{ref-Bolnick2003}{2003}); Butler, Sawyer, and Losos
(\protect\hyperlink{ref-Butler2007}{2007}); Bolnick and Lau
(\protect\hyperlink{ref-Bolnick2008}{2008}); Cooper, Gilman, and
Boughman (\protect\hyperlink{ref-Cooper2011}{2011}); but see Stuart et
al. (\protect\hyperlink{ref-Stuart2021}{2021}); Blain
(\protect\hyperlink{ref-Blain2022}{2022})). Accordingly, sexes should
also diverge in morphological traits associated with habitat use; such
ecologically-driven ``character displacement between the sexes'' is
therefore one explanation for phenotypic sexual dimorphism (S. P. De
Lisle and Rowe (\protect\hyperlink{ref-deLisleRowe2015}{2015}); S. Paiva
De Lisle S. P. and Rowe (\protect\hyperlink{ref-deLisleNco2018}{2018})).

We tested this link between release from predators and the evolution of
sexual dimorphism in a well-preserved, finely-resolved,
\textasciitilde16,000 year-long sequence of the fossil stickleback fish
(Gasterosteus doryssus). The fossil lineage appeared in the depositional
environment as a fully armored form, with complete pelvic girdles, two
pelvic spines, and three dorsal spines, on average (Michael A. Bell,
Travis, and Blouw (\protect\hyperlink{ref-Bell2006}{2006}); Stuart,
Travis, and Bell (\protect\hyperlink{ref-Stuart2020}{2020})). However,
this habitat appears to have been missing fish piscivores (e.g., trout
and other salmonids known to prey on modern stickleback; M. A. Bell
(\protect\hyperlink{ref-Bell2009}{2009})), relaxing putative selection
for armor (T. E. Reimchen (\protect\hyperlink{ref-Reimchen1980}{1980}),
T. E. Reimchen (\protect\hyperlink{ref-Reimchen1994}{1994}), O. Bell M.
A. and Koenings (\protect\hyperlink{ref-BellEtAl1993}{1993}); Bowne
(\protect\hyperlink{ref-Bowne1994}{1994}); Roesti and Schluter
(\protect\hyperlink{ref-Roestietal2023}{2023})). The lineage adaptively
evolved armor loss but also reduction in the means of several other
traits (Hunt, Bell, and Travis (\protect\hyperlink{ref-Hunt2008}{2008});
Stuart, Travis, and Bell (\protect\hyperlink{ref-Stuart2020}{2020});
Siddiqui et al.~in prep). In particular, tooth wear data suggest that
the population shifted to eat more planktonic prey through time,
indicative of increased open water habitat use (Purnell et al.~2007). If
this divergence in habitat use is partitioned between sexes (Reimchen T.
E. and Nosil (\protect\hyperlink{ref-ReimchenNosil2004}{2004});
Spoljaric and Reimchen
(\protect\hyperlink{ref-SpoljaricReimchen2008}{2008});D. S. Reimchen T.
E. (\protect\hyperlink{ref-Reimchen2016}{2016})), then we should expect
sexual dimorphism to evolve for traits related to habitat use in G.
doryssus.

To test this prediction, we imputed the sex of individual fossils, which
is unobserved in practice, with Multiple Imputation by Chained Equations
(MICE) (Buuren and Groothuis-Oudshoorn
(\protect\hyperlink{ref-MICE}{2011})), using observed data from extant
stickleback of known sex to train the imputation model. Then we tracked
the multivariate evolution of sexual dimorphism over
\textasciitilde16,000 years for 16 traits related to swimming, feeding
and defense.

\hypertarget{methods-and-materials}{%
\section{Methods and Materials}\label{methods-and-materials}}

A major challenge in this work is inferring the sex of fossil
stickleback specimens. Indeed, for most species, sex cannot be detected
directly, except for lineages whose sexes are distinguished by the
presence or absence of sex-specific characters (Hone and Mallon
(\protect\hyperlink{ref-Hone2017}{2017}); Saitta et al.
(\protect\hyperlink{ref-SaittaEtAl2020}{2020})). Instead,
paleobiologists resort to statistical detection of sex and sexual
dimorphism, including tests for bimodality in trait distributions (e.g.,
Hone and Mallon (\protect\hyperlink{ref-Hone2017}{2017})) and divergence
in growth curves (e.g., Saitta et al.
(\protect\hyperlink{ref-SaittaEtAl2020}{2020})). We used a third
approach to infer fossil sex and study the evolution of sexual
dimorphism by taking advantage of the fact that G. doryssus is part of
the extant threespine stickleback species complex, Gasterosteus
aculeatus. We measured the same trait set from a phenotypically diverse
set of extant G. aculeatus populations, for whom individual sex was
known. We then combined fossil and extant data sets, treated fossil sex
as missing data, and used the MICE (Buuren and Groothuis-Oudshoorn
(\protect\hyperlink{ref-MICE}{2011})) multiple imputation (Little and
Rubin (\protect\hyperlink{ref-little2002statistical}{2002})) algorithm,
and implemented in R (Team
(\protect\hyperlink{ref-R2022language}{2022})), to impute individual sex
of the fossil data set 100 times. We then fit a modified
Ornstein--Uhlenbeck (OU) (Uhlenbeck and Ornstein
(\protect\hyperlink{ref-OUProcess}{1930})) model using a Bayesian
framework to the data to test for the evolution of sexual dimorphism in
each trait.

@@

We used Gasterosteus doryssus data that were previously reported by
Stuart, Travis, and Bell (\protect\hyperlink{ref-Stuart2020}{2020}),
Voje, Bell, and Stuart (\protect\hyperlink{ref-Voje2022}{2022}), and
Siddiqui et al.~(in review). Briefly, the data were collected from
fossil Series K from Quarry D (Cerasoni et al., in review), dug from an
open pit diatomite mine at 9.526° N, 119.094° W, near Fernley, Nevada,
USA. Series K consisted of 18 samples taken at \textasciitilde1000-year
intervals, and mean sample times span \textasciitilde16,363 years. Fish
from series K were measured for 16 ecomorphological traits related to
armor, swimming, and feeding (Table 1). Series K started at a previously
documented horizon when a low armored lineage of stickleback with zero
to one dorsal spines, zero pelvic spines, and highly reduced pelvises
was completely replaced by a high armored lineage of stickleback with
three dorsal spines, two pelvic spines, and complete (Bell et al.~2006;
Bell 2009; Stuart et al.~2020). This lineage subsequently evolved
reduction in armor, body size, and traits related to swimming and
feeding (Bell et al.~2006; Stuart et al.~2020; Siddiqui et al.~in
review). The tempo and mode of armor reduction during this sequence
suggests adaptive evolution by natural selection (Hunt et al.~2008), and
we focus on the multivariate evolution of sexual dimorphism by this
second lineage.

The fossil data consists of 814 specimens with unknown sex over 18 time
periods spaced \textasciitilde1000 years apart. Figure 1 shows the
sample size at each of the 18 time periods. There are at least 22
specimens at each time period with a high of 67 specimens in period 7.

@@

\caption{Fossil sampel size for each time period}\label{fig:unnamed-chunk-2}
\end{figure}

@@

To span the gamut of stickleback diversity for our predictive model, we
sampled modern stickleback from lakes containing generalist stickleback
populations (Hendry et al. (\protect\hyperlink{ref-Hendry2009}{2009});
Bolnick (\protect\hyperlink{ref-Bolnick2011}{2011})) and from lakes
containing benthic-limnetic species pairs (Baumgartner et al.~1988;
Schluter and McPhail (\protect\hyperlink{ref-Schluter1992}{1992}))
(Table 2 (What is this?)). The generalist populations were collected by
YES in 2013 and previously described in Stuart et al.
(\protect\hyperlink{ref-Stuart2017}{2017}). These samples were fixed in
formalin, then stained for bone with Alizarin Red in 2013. Benthic and
limnetic specimens were kindly loaned by D. Schluter and his lab at
University of British Columbia. They collected benthic and limnetic
individuals from Enos Lake in 1988 and from Emily Lake, Little Quarry
Lake, Paxton Lake, and Priest Lake in 2018. The Enos specimens had been
fixed whole in formalin and stored in 40\% isopropanol. The specimens
from the other lakes were initially preserved whole in 95\% ethanol in
the field before being gradually transferred to water then formalin in
the lab and ultimately stored in 40\% isopropanol. In 2019, we stained
these specimens for bone using Alizarin Red.

We next replicated fossil data collection (Table 1) on these extant
specimens. Standard length as well as pelvic-spine length on each side
were measured with calipers. We used a dissection microscope to count
dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays. Right
and left-side pelvic girdle lengths and ectocoracoid lengths were
measured from ventral photographs taken using a Canon EOS Rebel T7 with
a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy
stand. Specimens were held in place for ventral photographs using a
small tabletop vise with an attached scale bar. Lateral X-rays were used
to measure dorsal spine length, number of pterygiophores anterior to the
pterygiophore holding the third spine, length of the pterygiophore just
anterior to the third spine, cleithrum length, and pre-maxilla ascending
branch length. We also counted vertebrae from the X-rays: abdominal
vertebrae were counted anterior to the first vertebra with a haemal
spine contacting an anal fin pterygiophore. Caudal vertebrae were
posterior, including the first vertebra with the haemal spine contacting
the anal fin pterygiophore (following Aguirre, Walker, and Gideon
(\protect\hyperlink{ref-Aguirre2014}{2014}). X-rays were taken with an
AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the Field
Museum of Natural History. Specimens were exposed at 35kV and 4mA. Small
fish were exposed for 7s, medium fish for 8s, and large fish for 10s. We
developed the film and scanned individual images of each fish using the
B\&W Negatives setting on an Epson Perfection 4990 Photo flatbed at 2400
dpi. Measurements from photographs and X-rays were taken with FIJI
(Schindelin et al.~2012) and its plugin ObjectJ
(\url{https://sils.fnwi.uva.nl/bcb/objectj/}). All photographs, X-rays,
and ObjectJ files have been uploaded to Morphosource.org (accession \#
TBD). We dissected individuals from the generalist populations to
determine sex from the gonads. Individuals from the species-pair lakes
were sexed by Schluter and his lab, using a genotyping protocol (confirm
and cite).

The extant data used here consists of a total of 367 specimens all with
known sex. Of these, there are 202 and 165 female and male specimens,
respectively.

\hypertarget{outlier-analysis.}{%
\subsection{Outlier analysis.}\label{outlier-analysis.}}

To check for outliers, we calculated within-group means and standard
deviations for each trait separately for K series fossil specimens
(pooled across samples) and for extant specimens (pooled across lakes).
We noted trait values greater than 3.5 standard deviations from the mean
as potential outliers. We checked whether these potential outliers were
a result of data entry and collection error and corrected them if they
were. We turned the remaining outlier trait values to NAs. (confirm)
(Wait, what? Why? Anything that was 3.5 SD above the mean was recorded
as missing? How do you justify this? )

Missing data imputation, including fossil sex

Quantification and evolution of sexual dimorphism

What covariates do we have in the data: length, what else,

@@

Let \(\boldsymbol{W}\) be an \((n_{extant} + n_{fossil}) \times 1\)
vector of the covariate gender of the stickleback fish and
\(\boldsymbol{Y}\) be an \((n_{extant} + n_{fossil}) \times K\) matrix
of the \(K\) phenotypes of interest. Because the gender of the
fossilized stickleback fish is unobservable, we further define
\(\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T\)
where \(\boldsymbol{W}_{extant}\) and \(\boldsymbol{W}_{fossil}\) are
the \(n_{extant} \times 1\) and \(n_{fossil} \times 1\) vectors of the
observed extant gender and missing fossil gender, respectively.

We impute the missing gender for the fossil data by sampling from the
posterior predictive distribution
\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})\)
using the multiple imputation by chained equations (MICE) algorithm
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
with predictive mean matching. Traditionally, the choice for the number
of completed dats sets is a relatively small number such as \(M = 5\) or
\(M = 10\). However, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a larger number
of imputed data sets if the data users intend on performing Bayesians
analysis after imputation, which in this case, we do. Therefore, the
imputation algorithm is run to obtain a total of \(M = 100\) completed
datasets. In addition to this, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}) suggests rather than
using Rubin's combining rules to combine across imputed data sets,
instead pool all of the draws from the posterior distributions across
all of the imputed data sets to estimate the posterior distributions of
parameters of interest. We proceed with our Bayesian analysis in this
manner.

@@

As previously described, four models for the mean process were
considered for each trait. Within each trait, the model with the lowest
DIC was chosen for that train. All subsequent results presented here use
the model with the lowest DIC individually by trait. Table
\ref{whichmodel} shows the model that was selected for each trait. About
half of the traits had models with a trend, and most models did not
include an OU component. The traits that did include an OU component
were caudal vertebra (mcv), pelvic spine lentgh (lpt), abdominal
vertebra (mav), and standard length (stl) with caudal vertebra (mcv)
being the only model with both a trend and an OU component. This implies
that not only does caudal vertebra size have a linear trend across
gender, but the average size at one time period is highly correlated
with the average size at the next time period.

>>>>

Ecological release theory suggests that a population's niche should
change when important species interactions (e.g., resource competition)
are relaxed or removed (reviewed in Herrmann, Stroud, and Losos
(\protect\hyperlink{ref-Herrmannetal2021}{2021})). Removal is posited to
create ecological opportunities---i.e., aspects of the niche become
newly accessible, and the focal population shifts and/or expands its
resource use to the new opportunity (Parent and Crespi
(\protect\hyperlink{ref-ParentandCrespi2009}{2009}); Herrmann, Stroud,
and Losos (\protect\hyperlink{ref-Herrmannetal2021}{2021})). Ecological
release may then be followed by adaptive morphological evolution (Parent
and Crespi (\protect\hyperlink{ref-ParentandCrespi2009}{2009});
Herrmann, Stroud, and Losos
(\protect\hyperlink{ref-Herrmannetal2021}{2021})). For example, a
population undergoing ecological release can experience disruptive
selection on males and females stemming from intraspecific competition
over newly accessible resources (Bolnick and Doebeli
(\protect\hyperlink{ref-BolnickandDoebeli2003}{2003}); Bolnick and Lau
(\protect\hyperlink{ref-BolnickandLau2008}{2008}); Cooper, Gilman, and
Boughman (\protect\hyperlink{ref-Cooperetal2011}{2011})). This is
predicted to result in intersexual divergence in habitat use and
associated phenotypes (Schoener
(\protect\hyperlink{ref-Schoener1968}{1968}); Shine
(\protect\hyperlink{ref-Shine1989}{1989}); Bolnick and Doebeli
(\protect\hyperlink{ref-BolnickandDoebeli2003}{2003}); Butler and Losos
(\protect\hyperlink{ref-Butleretal2007}{2007}); Bolnick and Lau
(\protect\hyperlink{ref-BolnickandLau2008}{2008}); Cooper, Gilman, and
Boughman (\protect\hyperlink{ref-Cooperetal2011}{2011}); but see Stuart
et al. (\protect\hyperlink{ref-Stuartetal2021}{2021}); Blain
(\protect\hyperlink{ref-Blain2022}{2022})). Such competition-driven
``intraspecific character displacement'' between the sexes is therefore
one explanation for the evolution of sexual dimorphism (Pfennig and
Pfennig (\protect\hyperlink{ref-PfennigandPfennig2012}{2012}); De Lisle
and Rowe (\protect\hyperlink{ref-DeLisleandRowe2015}{2015}); De Lisle
and Rowe (\protect\hyperlink{ref-DeLisleandRowe2017}{2017}), De Lisle,
Paiva, and Rowe (\protect\hyperlink{ref-DeLisleetal2018}{2018})).

Much of the theory and empirical data for character displacement between
the sexes is based on release from resource competition specifically
(e.g., Bolnick and Doebeli
(\protect\hyperlink{ref-BolnickandDoebeli2003}{2003}); Cooper, Gilman,
and Boughman (\protect\hyperlink{ref-Cooperetal2011}{2011}); Pfennig and
Pfennig (\protect\hyperlink{ref-PfennigandPfennig2012}{2012}); De Lisle
and Rowe (\protect\hyperlink{ref-DeLisleandRowe2015}{2015}); De Lisle,
Paiva, and Rowe (\protect\hyperlink{ref-DeLisleetal2018}{2018})).
However, release from predation should also result in ecological release
followed by the evolution of increased sexual dimorphism because the
absence of predators should generate ecological opportunity (Reimchen
and Nosil (\protect\hyperlink{ref-ReimchenandNosil2004}{2004}); Parent
and Crespi (\protect\hyperlink{ref-ParentandCrespi2009}{2009});
Herrmann, Stroud, and Losos
(\protect\hyperlink{ref-Herrmannetal2021}{2021})).

Here, we tested the prediction that a release from predation should
result in the evolution of increased sexual dimorphism using a
well-preserved, finely-resolved sequence of a fossil threespine
stickleback fish (Gasterosteus doryssus). The sequence in this
depositional environment is comprised of two lineages (Bell
(\protect\hyperlink{ref-Bell2009}{2009}); Cerasoni, Bell, and Stuart
(\protect\hyperlink{ref-Cerasonietal2024}{2024})). Lineage I was a
low-armor form with zero to one dorsal spines and highly reduced
pelvises, on average. It lasted for at least 93,000 years before it was
replaced suddenly by a second lineage (Bell, Baumgartner, and Oslon
(\protect\hyperlink{ref-Belletal1985}{1985})). Lineage II appeared in
the depositional environment fully armored, with complete pelvic
girdles, two pelvic spines, and three dorsal spines, on average (Bell,
Travis, and Blouw (\protect\hyperlink{ref-Belletal2006}{2006}); Stuart,
Travis, and Bell (\protect\hyperlink{ref-Stuartetal2020}{2020})). It
then immediately began evolving adaptive reduction in its armor traits
(Hunt, Bell, and Travis (\protect\hyperlink{ref-Huntetal2008}{2008});
Stuart, Travis, and Bell (\protect\hyperlink{ref-Stuartetal2020}{2020}))
until it reached the same low-armor state previously held by lineage I.
The observation of 93,000 years of low-armor stasis in lineage I (Bell,
Baumgartner, and Oslon (\protect\hyperlink{ref-Belletal1985}{1985})) and
the observation of rapid evolution of armor loss by lineage II, both
suggest that this depositional environment lacked piscivorous fish
(e.g., trout and other salmonids known to prey on modern threespine
stickleback). Armor maintenance in extant threespine stickleback
(Gasterosteus aculeatus) correlates strongly the presence of vertebrate
piscivores; populations with less predation pressure typically have less
armor (Reimchen (\protect\hyperlink{ref-Reimchen1994}{1994}), Reimchen
and Nosil (\protect\hyperlink{ref-ReimchenandNosil2004}{2004}); Bell et
al. (\protect\hyperlink{ref-Belletal1993}{1993}); Roesti et al.
(\protect\hyperlink{ref-Roestietal2023}{2023})). Consistently, only
three fossil trout have been recorded from this paleolake basin, sampled
from the same rock that has revealed \textgreater15,000 threespine
stickleback fossils as well as occasional killifish (Fundulus
nevadensis) (Bell (\protect\hyperlink{ref-Bell2009}{2009}); Cerasoni,
Bell, and Stuart (\protect\hyperlink{ref-Cerasonietal2024}{2024})).
Finally, tooth wear data from the lineage II sequence suggest that
individuals in this lineage began eating more planktonic prey over time,
expanding toward an open-water niche from the benthic, bottom-feeding
niche they started with (Purnell et al.~2007). That this expansion into
open water by lineage II coincided with armor reduction is further
indicative of a limnetic system with fewer salmonid piscivores (Schluter
and McPhail (\protect\hyperlink{ref-SchluterandMcPhail1992}{1992});
Vamosi and Schluter
(\protect\hyperlink{ref-VamosiandSchluter2004}{2004}); Roesti et al.
(\protect\hyperlink{ref-Roestietal2023}{2023})). In sum, we infer that
lineage II likely started in a nearby paleolake basin that had predators
since it was armored when it arrived (Bell
(\protect\hyperlink{ref-Bell2009}{2009}); Cerasoni, Bell, and Stuart
(\protect\hyperlink{ref-Cerasonietal2024}{2024})). Lineage II then
experienced release from predators and dietary expansion in the focal
paleolake basin, generating a scenario consistent with the starting
conditions for models predicting the evolution of increased sexual
dimorphism following ecological release.

A major challenge in the study of fossilized sexual dimorphism is
assigning sex to individual specimens in the first place (Hone and
Mallon (\protect\hyperlink{ref-HoneandMallon2017}{2017}); Mallon
(\protect\hyperlink{ref-Mallon2017}{2017}); Saitta et al.
(\protect\hyperlink{ref-Saittaetal2020}{2020})). This typically cannot
be done directly, except for taxa whose sexes are distinguished by the
presence or absence of sex-specific characters that preserve well.
Instead, paleobiologists often resort to statistical detection of sex
and sexual dimorphism, including tests for normality and bimodality in
trait distributions (e.g., Mallon
(\protect\hyperlink{ref-Mallon2017}{2017})), mixture modeling (e.g.,
Mallon (\protect\hyperlink{ref-Mallon2017}{2017})), divergence in growth
curves (e.g., Saitta et al.
(\protect\hyperlink{ref-Saittaetal2020}{2020})), and emphasis on effect
size statistics rather than significance testing (e.g., Saitta et al.
(\protect\hyperlink{ref-Saittaetal2020}{2020})). However, dimorphic
signal can be masked by noise introduced by factors both biological and
artifactual, including extended growth during ontogeny, age-based bias
in survivorship, small sample sizes, time averaging, and taphonomic bias
(Godfrey, Lyon, and Sutherland
(\protect\hyperlink{ref-Godfreyetal1993}{1993}); Koscinski and
Pietraszewski
(\protect\hyperlink{ref-KoscinskiandPietraszewski2004}{2004}); Hone and
Mallon (\protect\hyperlink{ref-HoneandMallon2017}{2017}); reviewed in
Mallon (\protect\hyperlink{ref-Mallon2017}{2017}); reviewed in Saitta et
al. (\protect\hyperlink{ref-Saittaetal2020}{2020})). Thus, the best
approach to sex classification is likely one of total evidence (Saitta
et al. (\protect\hyperlink{ref-Saittaetal2020}{2020})), including
comparison to closely-related, extant species of known sex (e.g., Hone
and Mallon (\protect\hyperlink{ref-HoneandMallon2017}{2017}); Saitta et
al. (\protect\hyperlink{ref-Saittaetal2020}{2020})).

For our study, we inferred sex in G. doryssus fossils by comparing
multivariate morphological trait data from lineage II samples to the
same multivariate trait set collected from multiple populations in the
closely-related, extant Threespine Stickleback species complex
(Gasterosteus aculeatus). Crucially, we determined G. aculeatus sex
directly via dissection and/or PCR genotyping. This enabled us to use
Multiple Imputation by Chained Equations (MICE; Buuren and
Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011})) to build a
predictive multiple imputation algorithm (Little and Rubin
(\protect\hyperlink{ref-little2002statistical}{2002})) based on the
multivariate morphology of G. aculeatus individuals of known sex. We
applied this algorithm to the fossil data to impute individual sex based
on morphology, treating fossil sex as a missing variable. With sex
assigned to fossil specimens (and accounting for uncertainty in those
assignments), we fit a modified Ornstein-Uhlenbeck (OU) (Uhlenbeck and
Ornstein (\protect\hyperlink{ref-OUProcess}{1930})) model using a
Bayesian framework to test for evolution of sexual dimorphism in each
trait over \textasciitilde16,000 years of lineage II. We hypothesized
that sexual dimorphism should increase following ecological release from
predators.

@@

We used Gasterosteus doryssus data that were previously reported by
Stuart, Travis, and Bell (\protect\hyperlink{ref-Stuartetal2020}{2020}),
Voje, Bell, and Stuart (\protect\hyperlink{ref-Vojeetal2022}{2022}), and
Siddiqui et al. (\protect\hyperlink{ref-Siddiquietal2024}{2024}).
Briefly, the data were collected from fossil Series K from Quarry D
(Cerasoni, Bell, and Stuart
(\protect\hyperlink{ref-Cerasonietal2024}{2024})), dug from an open pit
diatomite mine at 9.526° N, 119.094° W, near Fernley, Nevada, USA.
Series K consisted of 18 samples taken at \textasciitilde1000-year
intervals, and mean sample times span \textasciitilde16,363 years. Fish
from series K were measured for 16 ecomorphological traits related to
armor, swimming, and feeding (Table 1). Series K started at a previously
documented horizon when a low armored lineage of stickleback with zero
to one dorsal spines, zero pelvic spines, and highly reduced pelvises
(i.e., lineage I described above) was completely replaced by a high
armored lineage of stickleback with three dorsal spines, two pelvic
spines, and complete pelvises (i.e., lineage II described above) (Bell,
Travis, and Blouw (\protect\hyperlink{ref-Belletal2006}{2006}); Bell
(\protect\hyperlink{ref-Bell2009}{2009}); Stuart, Travis, and Bell
(\protect\hyperlink{ref-Stuartetal2020}{2020})). Lineage II subsequently
evolved reduction in armor, body size, and traits related to swimming
and feeding (Bell, Travis, and Blouw
(\protect\hyperlink{ref-Belletal2006}{2006}); Stuart, Travis, and Bell
(\protect\hyperlink{ref-Stuartetal2020}{2020}); Siddiqui et al.
(\protect\hyperlink{ref-Siddiquietal2024}{2024})). The tempo and mode of
armor reduction during this sequence suggests adaptive evolution by
natural selection (Hunt, Bell, and Travis
(\protect\hyperlink{ref-Huntetal2008}{2008})), and we focus on the
multivariate evolution of sexual dimorphism by this second lineage.

The lineage II fossil data consist of 814 specimens of unknown sex
sampled across 18 time periods spaced \textasciitilde1000 years apart.
Figure 1 shows the sample size for each of the 18 samples. There are at
least 22 specimens in each sample with a high of 67 specimens in sample
7.

@@

\caption{Fossil sample size for each time period}\label{fig:unnamed-chunk-2}
\end{figure}

@@

To span the gamut of stickleback diversity for our predictive imputation
model, we sampled modern stickleback from lakes containing generalist
stickleback populations (Hendry et al.
(\protect\hyperlink{ref-Hendryetal2009}{2009}); Bolnick
(\protect\hyperlink{ref-Bolnick2011}{2011})) and from lakes containing
benthic-limnetic species pairs (Baumgartner, Bell, and Weinberg
(\protect\hyperlink{ref-Baumgartneretal1988}{1988}); Schluter and
McPhail (\protect\hyperlink{ref-SchluterandMcPhail1992}{1992})). The
generalist populations used here were the remainder of collections made
by YES in 2013 and were previously described in Stuart et al.
(\protect\hyperlink{ref-Stuartetal2017}{2017}) (Table S1). These samples
were fixed in formalin, then stained for bone with Alizarin Red in 2013.
Benthic and limnetic specimens were kindly loaned by D. Schluter and his
lab at University of British Columbia. They collected benthic and
limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little
Quarry Lake, Paxton Lake, and Priest Lake in 2018 (Table S1). The Enos
specimens had been fixed whole in formalin and stored in 40\%
isopropanol. The specimens from the other lakes were initially preserved
whole in 95\% ethanol in the field before being gradually transferred to
water then formalin in the lab and ultimately stored in 40\%
isopropanol. In 2019, we stained these specimens for bone using Alizarin
Red. We next replicated fossil data collection (Table 1) on these extant
specimens. Standard length as well as pelvic-spine length on each side
were measured with calipers. We used a dissection microscope to count
dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays. Right
and left-side pelvic girdle lengths and ectocoracoid lengths were
measured from ventral photographs taken using a Canon EOS Rebel T7 with
a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy
stand. Specimens were held in place for ventral photographs using a
small tabletop vise with an attached scale bar. Lateral X-rays were used
to measure dorsal spine length, number of pterygiophores anterior to the
pterygiophore holding the third spine, length of the pterygiophore just
anterior to the third spine, cleithrum length, and pre-maxilla ascending
branch length. We also counted vertebrae from the X-rays: abdominal
vertebrae were counted anterior to the first vertebra with a haemal
spine contacting an anal fin pterygiophore. Caudal vertebrae were
posterior, including the first vertebra with the haemal spine contacting
the anal fin pterygiophore (following Aguirre, Walker, and Gideon
(\protect\hyperlink{ref-Aguirreetal2014}{2014})). X-rays were taken with
an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the
Field Museum of Natural History. Specimens were exposed at 35kV and 4mA.
Small fish were exposed for 7s, medium fish for 8s, and large fish for
10s. We developed the film and scanned individual images of each fish
using the B\&W Negatives setting on an Epson Perfection 4990 Photo
flatbed at 2400 dpi. Measurements from photographs and X-rays were taken
with FIJI (Schindelin et al.
(\protect\hyperlink{ref-Schindelinetal2012}{2012})) and its plugin
ObjectJ (\url{https://sils.fnwi.uva.nl/bcb/objectj/}).

We dissected individuals from the generalist populations (Table S1) to
determine sex from the gonads. Individuals from the species-pair lakes
(TableS1) were previously sexed by the Schluter lab, using either
dissection or a genotyping protocol (Whom, personal communication).

The extant data used here consists of a total of 367 specimens all with
known sex (Table S1). Of these, there are 202 and 165 female and male
specimens, respectively.

\hypertarget{outlier-analysis-and-size-correction.}{%
\subsection{Outlier analysis and size
correction.}\label{outlier-analysis-and-size-correction.}}

To check for outliers, we calculated within-group means and standard
deviations for each trait separately for K series fossil specimens
(pooled across samples) and for extant specimens (within generalist,
benthic, or limnetic categories). We noted trait values greater than 3.0
standard deviations from the mean as potential outliers. We deemed 3.0
s.d. to be a reasonable threshold for detecting errors without excluding
biologically relevant values. We checked whether these potential
outliers were a result of data entry and collection error and corrected
them if they were. We turned the remaining outlier trait values to NAs.
We size-corrected continuous traits only, as they varied with size,
unlike count traits that are fixed during early development. We
regressed each trait on standard length using a mixed-model regression,
pooling all specimens, following Stuart et al.
(\protect\hyperlink{ref-Stuartetal2017}{2017}). We appended the size
corrected data to the uncorrected trait data frame and used all of our
data to build the imputation model described next.

\hypertarget{missing-data-imputation-including-fossil-sex}{%
\subsection{Missing data imputation, including fossil
sex}\label{missing-data-imputation-including-fossil-sex}}

We used multiple imputation (Little and Rubin
(\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
of the fossils many times. Imputations were performed using the MICE
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
algorithm and implemented in R (Team
(\protect\hyperlink{ref-R2022language}{2022})) with M = 100 imputations
performed. After imputing the sex of the fossil data, each completed
data set is then used to fit a modified Ornstein--Uhlenbeck (OU)
(Uhlenbeck and Ornstein (\protect\hyperlink{ref-OUProcess}{1930})) model
using a Bayesian framework to look for sexual dimorphism across a
variety of stickleback phenotypes.

@@

Let \(\boldsymbol{W}\) be an \((n_{extant} + n_{fossil}) \times 1\)
vector of the covariate sex of the stickleback fish and
\(\boldsymbol{Y}\) be an \((n_{extant} + n_{fossil}) \times K\) matrix
of the \(K\) phenotypes of interest. Because the sex of the fossilized
stickleback fish is unobservable, we further define
\(\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T\)
where \(\boldsymbol{W}_{extant}\) and \(\boldsymbol{W}_{fossil}\) are
the \(n_{extant} \times 1\) and \(n_{fossil} \times 1\) vectors of the
observed extant sex and missing fossil sex, respectively.

We impute the missing sex for the fossil data by sampling from the
posterior predictive distribution
\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})\)
using the multiple imputation by chained equations (MICE) algorithm
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
with predictive mean matching. Traditionally, the choice for the number
of completed dats sets is a relatively small number such as \(M = 5\) or
\(M = 10\). However, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a larger number
of imputed data sets if the data users intend on performing Bayesians
analysis after imputation, which in this case, we do. Therefore, the
imputation algorithm is run to obtain a total of \(M = 100\) completed
datasets. In addition to this, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}) suggests rather than
using Rubin's combining rules to combine across imputed data sets,
instead pool all of the draws from the posterior distributions across
all of the imputed data sets to estimate the posterior distributions of
parameters of interest. We proceed with our Bayesian analysis in this
manner.

@@

As previously described, four models for the mean process were
considered for each trait. Within each trait, the model with the lowest
DIC was chosen for that train. All subsequent results presented here use
the model with the lowest DIC individually by trait. Table
\ref{whichmodel} shows the model that was selected for each trait. About
half of the traits had models with a trend, and most models did not
include an OU component. The traits that did include an OU component
were caudal vertebra (mcv), pelvic spine lentgh (lpt), abdominal
vertebra (mav), and standard length (stl) with caudal vertebra (mcv)
being the only model with both a trend and an OU component. This implies
that not only does caudal vertebra size have a linear trend across sex,
but the average size at one time period is highly correlated with the
average size at the next time period.
